---
title: "Citizen science data exploration"
author: "Ben Goldstein"
date: "1/28/2020"
output: html_document
---


This script reads in eBird data for the four target species. 
It will only work on Ben's computer as he has the data.

```{r, include = FALSE}
library(tidyverse)
library(XML)
```

Our four species are: verdin, black-tailed gnatcatcher, crissal thrasher, and least Bell's vireo.

```{r}
target_common_names <- c("Verdin", "Black-tailed_Gnatcatcher", "Crissal_Thrasher", "Bells_Vireo")
```


First, read in the data we have. Data come from 4 sources: eBird, iNaturalist, GBIF, and NABBS.

```{r message = F}
# eBird data
ebird_CA_obs <- read_csv("../eBird_Data/subsets-2019/state_rdbs/CA_Aug2019_species_counts.csv") %>% 
                  filter(name_clean %in% target_common_names)
ebird_CA_checklists <- read_csv("../eBird_Data/subsets-2019/state_rdbs/CA_Aug2019_checklist_info.csv")


ebird_NV_obs <- read_csv("../eBird_Data/subsets-2019/state_rdbs/NV_Aug2019_species_counts.csv") %>% 
                  filter(name_clean %in% target_common_names)
ebird_NV_checklists <- read_csv("../eBird_Data/subsets-2019/state_rdbs/NV_Aug2019_checklist_info.csv")


ebird_AZ_obs <- read_csv("../eBird_Data/subsets-2019/state_rdbs/AZ_Aug2019_species_counts.csv") %>% 
                  filter(name_clean %in% target_common_names)
ebird_AZ_checklists <- read_csv("../eBird_Data/subsets-2019/state_rdbs/AZ_Aug2019_checklist_info.csv")


ebird_UT_obs <- read_csv("../eBird_Data/subsets-2019/state_rdbs/UT_Aug2019_species_counts.csv") %>% 
                  filter(name_clean %in% target_common_names)
ebird_UT_checklists <- read_csv("../eBird_Data/subsets-2019/state_rdbs/UT_Aug2019_checklist_info.csv")


# 
species_names <- ebird_CA_obs %>% 
                 select(name_clean, SCIENTIFIC.NAME) %>% 
                 distinct(name_clean, SCIENTIFIC.NAME)

# North American Breeding Bird Survey data
# From https://www.pwrc.usgs.gov/bbs/
NABBS_files <- list.files("NABBS_data_28012020", full.names = T)
NABBS_data <- lapply(NABBS_files, read_csv)

# Global Biodiversity Information Facility data
# From https://doi.org/10.15468/dl.r49uys (csv from https://doi.org/10.15468/dl.qfmedi)
gbif_data <- read_delim(file = "gbif_data_28012020/0000808-200127171203522.csv", 
                        delim = "\t", quote = "")
```


Let's explore the data a bit. What do we have?

### eBird exploration

We have 33655 instances of positive IDs of our four species in NV, and 75737 in CA. But how many of each species?

```{r}
ebird_CA_obs %>% 
  count(name_clean) %>% 
  arrange(-n)
ebird_NV_obs %>% 
  count(name_clean) %>% 
  arrange(-n)
ebird_AZ_obs %>% 
  count(name_clean) %>% 
  arrange(-n)
ebird_UT_obs %>% 
  count(name_clean) %>% 
  arrange(-n)
```

What's the spatial distribution of the data?

```{r}
ebird_NV_obs_w_info <- left_join(ebird_NV_obs, ebird_NV_checklists)
ebird_CA_obs_w_info <- left_join(ebird_CA_obs, ebird_CA_checklists)
ebird_AZ_obs_w_info <- left_join(ebird_AZ_obs, ebird_AZ_checklists)
ebird_UT_obs_w_info <- left_join(ebird_UT_obs, ebird_UT_checklists)

ebird_obs_w_info <- bind_rows(ebird_NV_obs_w_info, 
                              ebird_CA_obs_w_info, 
                              ebird_AZ_obs_w_info,
                              ebird_UT_obs_w_info)

us <- map_data("state", region = c("Nevada", "California", "Arizona", "Utah"))

ggplot() +
  geom_map(data = us, map = us, aes(long, lat, map_id = region), 
           col = "black", fill = "white") +
  geom_point(data = ebird_obs_w_info, 
             aes(LONGITUDE, LATITUDE, col = name_clean), cex = 0.2) + coord_fixed()
```

